---
title: "Edx MovieLens Project"
date: "June 6, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Introduction
The goal of this project is to create a model that accurately predicts moving ratings for users. The data used in this project is from the MovieLens data set. The performance metric used was RMSE.

Key steps in the project were:

1. Data exploration and cleansing

2. Feature engineering

3. Model training

4. Model comparison and test estimation.

(Additional note: This data set large. This experiment was performed on a workstation with 16gb of ram. Workstations with less than 16gb of ram may not be able to process the data set.)

# Anlysis

## Data Load and libraries

First step was to load the data and libraries.

```{r}
#Load MovieLens dataset (Code copied from Edx project instructions)

#load packages

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data

set.seed(1) # if using R 3.6.0: set.seed(1, sample.kind = "Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

```{r}
#load libraries

library(tidyverse)
library(splitstackshape)
```

## Data cleansing and error checking

Next we checked the data for errors.

```{r}
#check for NA's

sum(is.na(edx)) 

#view summaries of each field for 0's, outliers, and anomalies

summary(edx$userId)
summary(edx$rating)
summary(edx$movieId)
data.frame(table(edx$genres))

```

After reviewing the data there seem to be no data integrity or outline problems. There were 7 movies that did not have a genres listed. Genres also contains multiple entries. These will be broken out into individual dummy variables. Looking at the data, additional features could be useful for the mode. We will engineer features for average movie rating, individuals average movie rating, and individuals average movie ratings for the most popular genres.

## Feature engineering

Next we engineered additional features for the data set.

```{r, cache.lazy=FALSE}
#create dummy variables for genres

genres <- cSplit(edx[6], 'genres', sep="|", type.convert=FALSE)
genres <- unique(genres[,1])

for (i in 1:19) {
  gen <- genres[i]
  index <- grep(gen, edx$genres)
  edx[index,6+i] <-1
  edx[-index,6+i] <-0
}

for (i in 1:19) {
  gen <- genres[i]
  index <- grep(gen, validation$genres)
  validation[index,6+i] <-1
  validation[-index,6+i] <-0
}

#create average movie ratings and user ratings

avgmovierating <- edx %>% group_by(movieId) %>% summarize(avgmovierating = mean(rating))
avguserrating <- edx %>% group_by(userId) %>% summarize(avguserrating = mean(rating))

edx <- merge(edx, avgmovierating, 'movieId', all.x = TRUE)
edx <- merge(edx, avguserrating, 'userId', all.x = TRUE)

validation <- merge(validation, avgmovierating, 'movieId', all.x = TRUE)
validation <- merge(validation, avguserrating, 'userId', all.x = TRUE)

#create individual average movie ratings for top 5 genreas

avguserrating_comedy <- edx[edx$V7 == 1,] %>% group_by(userId) %>% summarize(avguserrating_comedy = mean(rating))
edx <- merge(edx, avguserrating_comedy, 'userId', all.x = TRUE)
validation <- merge(validation, avguserrating_comedy, 'userId', all.x = TRUE)

avguserrating_action <- edx[edx$V8 == 1,] %>% group_by(userId) %>% summarize(avguserrating_action = mean(rating))
edx <- merge(edx, avguserrating_action, 'userId', all.x = TRUE)
validation <- merge(validation, avguserrating_action, 'userId', all.x = TRUE)

avguserrating_adventure <- edx[edx$V10 == 1,] %>% group_by(userId) %>% summarize(avguserrating_adventure = mean(rating))
edx <- merge(edx, avguserrating_adventure, 'userId', all.x = TRUE)
validation <- merge(validation, avguserrating_adventure, 'userId', all.x = TRUE)

avguserrating_drama <- edx[edx$V12 == 1,] %>% group_by(userId) %>% summarize(avguserrating_drama = mean(rating))
edx <- merge(edx, avguserrating_drama, 'userId', all.x = TRUE)
validation <- merge(validation, avguserrating_drama, 'userId', all.x = TRUE)

avguserrating_crime <- edx[edx$V13 == 1,] %>% group_by(userId) %>% summarize(avguserrating_crime = mean(rating))
edx <- merge(edx, avguserrating_crime, 'userId', all.x = TRUE)
validation <- merge(validation, avguserrating_crime, 'userId', all.x = TRUE)

#view summary of engineered features
summary(edx[,c(3,26:32)])

#remove extra tables and clear memory
rm(avgmovierating,avguserrating,avguserrating_action,avguserrating_adventure,avguserrating_comedy,avguserrating_crime, avguserrating_drama, gen, genres, i, index)
gc()
```

The engineered features contain NA's that need to be removed because some users have not rated those genres. The NAs will be replaced with the average genre rating for users who haven't rated that genre.

```{r, cache.lazy=FALSE}
#replace NA's with means
edx$avguserrating_comedy[is.na(edx$avguserrating_comedy)] <- mean(na.omit(edx$avguserrating_comedy))
validation$avguserrating_comedy[is.na(validation$avguserrating_comedy)] <- mean(na.omit(edx$avguserrating_comedy))

edx$avguserrating_action[is.na(edx$avguserrating_action)] <- mean(na.omit(edx$avguserrating_action))
validation$avguserrating_action[is.na(validation$avguserrating_action)] <- mean(na.omit(edx$avguserrating_action))

edx$avguserrating_adventure[is.na(edx$avguserrating_adventure)] <- mean(na.omit(edx$avguserrating_adventure)) 
validation$avguserrating_adventure[is.na(validation$avguserrating_adventure)] <- mean(na.omit(edx$avguserrating_adventure))

edx$avguserrating_drama[is.na(edx$avguserrating_drama)] <- mean(na.omit(edx$avguserrating_drama))
validation$avguserrating_drama[is.na(validation$avguserrating_drama)] <- mean(na.omit(edx$avguserrating_drama))

edx$avguserrating_crime[is.na(edx$avguserrating_crime)] <- mean(na.omit(edx$avguserrating_crime))
validation$avguserrating_crime[is.na(validation$avguserrating_crime)] <- mean(na.omit(edx$avguserrating_crime))

summary(edx[,c(3,26:32)])
```

NA's have been replaced and features are ready for modeling. 

## Modeling

We will attempt to fit a simple linear model.

```{r, cache.lazy=FALSE}
#fit LM model
fit.lm <- lm(rating~., data = edx[,c(3,7:32)])
summary(fit.lm)

#train RMSE
predict.lm <- predict(fit.lm, newdata = edx[,c(3,7:32)])
sqrt(mean((edx$rating - predict.lm)^2))

```

The majority of the features in the model are significant and the F-statistic is significant. The RMSE on the training set was .8698, within the target range of <= 0.87750. We will remove non-significant(p>.05) variables and refit the model.

```{r, cache.lazy=FALSE}
#remove previous fit to save memory
rm(fit.lm)
rm(predict.lm)

#fit LM model and remove non-signicant variables
fit.lm <- lm(rating~., data = edx[,c(3,7:16,18:21,23:24,26:32)])
summary(fit.lm)

#train RMSE
predict.lm <- predict(fit.lm, newdata = edx[,c(3,7:32)])
sqrt(mean((edx$rating - predict.lm)^2))
```

The model and remaining variables are significant. The RMSE on the training data was .8698, within the target range of <= 0.87750. This model should be suitable for our final model.

# Results

Lastly we tested the linear model against the validation set to provide a final estimate of performance.

```{r, cache.lazy=FALSE}
#remove previous predict to save memory
rm(predict.lm)

#final validation RMSE
predict.lm <- predict(fit.lm, newdata = validation[,c(3,7:32)])
sqrt(mean((validation$rating - predict.lm)^2))
```
The final model performance on the validation set was .8772, within the target range of <= 0.87750.

# Conclusion

The final linear model, with a RMSE of .8772, was within the target range of <= 0.87750. Additional features were engineered that ended up being significant to the model. These included variables for genres, average movie ratings, user's average ratings, and user's average ratings for genres.


